---
title: "CS-201 First Look at R"
author: "Shaylee Ferrer"
date: 8/26/26
format: html
editor: visual
toc: true
---

This is the first class session walking through code together!

We will be programming in R and in this class will primarily use `qmd` documents to keep things tidy and presentable. Each code chunk below is where you will run R code. Feel free to add notes as we go along.

## Navigating RStudio

By default, everyone should have the same four-pane layout:

1.  **Top left**: Source editor

2.  **Bottom left**: Console (and Terminal and Background Jobs)

3.  **Top Right**: Environment

4.  **Bottom Right**: Files, Plots, Help, etc.

### Customizing look & feel

Feel free to customize your RStudio.\
Go to Tools \> Global Options \> Appearance to select a color scheme you like.

I also personally like to enable these:

- Tools \> Global Options \> Code \> Insert matching parens/quotes

- Tools \> Global Options \> Console \> Show syntax highlighting in console input

I don't recommend rearranging your pane layout at this time.

## Running Code

Try running some code

### Load Library

```{r}
library(ggplot2)

```

### Some scripting

```{r}
#This is a comment
#Here we are assigning a value to a variable
quiz1 <- 88
quiz2 <- 94
quiz3 <- 79

quiz1

#getting the average
(quiz1 + quiz2 + quiz3) / 3

#getting the mean
mean(quiz1 + quiz2 + quiz3) #wrong cause your just adding them
mean(quiz1, quiz2, quiz3) #wrong cause it only returns the mean of the first argument
mean(c(quiz1, quiz2, quiz3))

#creatig a variable that holds info
list <- (c(quiz1, quiz2, quiz3))
mean(list) #makes it easier and less repetative

average <- mean(list)

quiz3 <- 99

average

list <- c(quiz1, quiz2, quiz3)
average <- mean(list)
round(average, 2)

```

### More Coding

```{r}
#ctr, opt, I makes a new code block
scores <- c(88, 94, 79, 91, 85)
scores

mean(scores)

min(scores) #minimum

max(scores) #maximum

length(scores) #number of things in the list

summary(scores)

students <- c("Alice", "Bob", "Carter", "Dylan", "Ethan")
students

#bar plot
barplot(scores, names.arg = students, ylim = c(0, 100), main = "Math Class", xlab = "Students", y = "Test Scores")
```

## Introducing `ggplot2`

#### Make a Dataframe

```{r}

grades_df <- data.frame(student = students, score = scores)

grades_df

summary(grades_df)

nrow(grades_df) #number of rows
ncol(grades_df) #number of collumns
dim(grades_df) #dimensions

```

#### Try ggplot

```{r}

ggplot(data = grades_df, aes(x = student, y = score))

ggplot(data = grades_df, aes(x = reorder(student, +score), y = score, group = 1)) +
  geom_col(fill = "navy") +
  geom_point() +
  geom_line() +
  theme_minimal() +
  labs(title = "Math Class",
       subtitle = "Quiz Chapter 1",
       x = "Students",
       y = "Scores")

```

#### Mutate the Dataframe

```{r}

library(tidyverse)

            #get this by hitting comand+shift+M 
grades_df <- grades_df |>
  mutate(study_hours = c(2, 5, 1, 4, 3))

```

#### Make a Scatterplot

```{r}

ggplot(data = grades_df, aes(x = study_hours, y = score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "lightblue") +
  geom_point(color = "navy") +
  theme_classic() +
  labs(title = "Study Observations",
       subtitle = "What do you see?",
       x = "Study Hours",
       y = "Student Scores") +
  geom_line(color = "navy")

  
#cannot use fill (use color instead), don't use reorder
  
```

## Extra Credit (1pt)

What result do you expect when you run this expression?

```{r}

1+1+1

quiz1+ quiz2

options(scipen = 999)

print(0.1, 20)
print(0.5, 20)

#why does it return that number?
0.1 + 0.1 + 0.1 - 0.3

```

What did you observe?

> Answer:

Why do you think this is happening? (You can look online for help)

> Answer:
