---
title: "Exploratory Data Analysis, part 2"
author: "Amber Camp"
date: "`r Sys.Date()`"
format: html
editor: visual
---

# Exploratory Data Analysis Part 2

Continuing on from Part 1. Start by loading libraries and reading in data, as needed

## Packages

```{r, message=FALSE}
#| label: packages

library(tidyverse)
library(janitor)
library(datasauRus)
```

## The Datasaurus Dozen

Last class, we found that all 13 `datasauRus` datasets have identical means, standard deviations, and correlations. Here's the table again:

```{r, message=FALSE}
#| label: datasauRus dozen descriptive

datasaurus_dozen |> 
  group_by(dataset) |> 
  summarise(
    n = n(),
    mean_x = mean(x),
    mean_y = mean(y),
    sd_x = sd(x),
    sd_y = sd(y),
    corr = cor(x, y)) |> 
  mutate(across(where(is.numeric), function(x) round(x, 2)))

```

What did you guess? Let's find out.

Plot the whole data frame by data set, facet wrapped.

```{r, message=FALSE}
#| label: plot datasauRus dozen by facets

ggplot(datasaurus_dozen, aes(x = x, y = y)) +
  geom_point(size = 0.9) +
  facet_wrap(~dataset, ncol = 5) +
  theme(legend.position = "none")

```

### Why this matters

These data sets were built by Justin Matejka and George Fitzmaurice (2017) as an extension of Anscombe's Quartet. The point is not that summary statistics are wrong–they're correct, and they're useful. The point is that [**summary statistics are a compression**]{.underline}, and every compression throws something away.

Summary statistics describe a distribution but do not uniquely determine it. Radically different data can produce identical numbers, so while you should always report a mean, sd, or correlation, never do so without also looking at the shape of the data. Plot first, or at least plot too.

**This is the whole justification for EDA. It is also why your assignment this week asks for summary stats *and* plots, rather than one or the other.**

## Data

Read in the Pokemon data and clean the column headers using janitor.

```{r, message=FALSE}
#| label: load pokedata

pokedata <- read_csv("data/Pokemon.csv")

# clean variable names with janitor
pokedata <- pokedata |>
  clean_names()

```

Remember we have some data that we want to treat as factor, so let's go ahead and convert them now. This is carried over from last class

```{r, message=FALSE, results='hide'}
#| label: clean data classes (from last class)

# skim_before <- skimr::skim(pokedata)

pokedata$number <- as.factor(pokedata$number)
pokedata$type_1 <- as.factor(pokedata$type_1)
pokedata$name <- as.factor(pokedata$name)
pokedata$type_2 <- as.factor(pokedata$type_2)
pokedata$generation <- as.factor(pokedata$generation)

# skim_after <- skimr::skim(pokedata)

```

By the way, if you want to save this as a new csv, just run the below. I'm going to leave it optional for now.

```{r}
#| label: write df to new csv

#write_csv(pokedata, "data/pokemon_cleaned.csv")

```

## Box plot time! (Extra credit)

Who's claiming the extra credit? Take us through it.

```{r, message=FALSE}
#| label: basic box plot

ggplot(pokedata, aes(y = hp)) +
  geom_boxplot()

```

How do we read box plots? Also called box and whisker plot/diagram.

- Middle line: Median

- Box: the middle 50% (IQR interquartile range)

- Whiskers: first and fourth quartiles

- Dots: outliers

### Now let's check that against the numbers

A box plot is not a new kind of information. It's the summary statistics from last class, drawn to scale. Let's compute the pieces and see whether they match what we just said.

```{r, message=FALSE}
#| label: hp descriptive summary

pokedata |> 
  summarise(
    min = min(hp),
    lower_fence = quantile(hp, 0.25) - 1.5 * IQR(hp),
    q1 = quantile(hp, 0.25),
    median = median(hp),
    q3 = quantile(hp, 0.75),
    iqr = IQR(hp),
    upper_fence = quantile(hp, 0.75) + 1.5 * IQR(hp),
    max = max(hp))

```

- **q1** is the 25th percentile: 25% of pokemon have `hp` below this value.

- **q3** is the 75th percentile: 75% of pokemon have `hp` below this value.

- **IQR** is `q3 - q1`, the range of the middle 50%.

- The **fences** are `q1 - 1.5 * IQR` and `q3 + 1.5 * IQR`. These are the conventional cutoffs for "unusual" outliers.

Match each number to a landmark on the plot above:

- Middle line: the **median** (the 50th percentile). Half the data is above, half below.

- Box: spans **q1 to q3**, so the box itself is the IQR — the middle 50% of the data.

- Whiskers: extend to the most extreme data point that is still **inside the fences**. They do *not* automatically go to the min and max. Compare `max` to `upper_fence` above and you'll see why.

- Dots: individual observations that fall **outside the fences**. These are the conventional "outliers."

Note what a boxplot does *not* show you: it has no idea whether your data is one clump, two clumps, or a dinosaur. Keep that in mind.

```{r, message=FALSE}
#| label: build various box plots

# build a box plot of type_1 and their hp
ggplot(data = pokedata, aes(x = type_1, y = hp)) +
  geom_boxplot()

# reorder type_1 by median hp
ggplot(data = pokedata, aes(x = fct_reorder(type_1, hp, median), y = hp)) +
  geom_boxplot()

# add individual points
ggplot(data = pokedata, aes(x = fct_reorder(type_1, hp, median), y = hp)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.3, width = 0.2, color = "purple") # alpha changes transparency, the smaller the more transparent

# facet wrap by generation + rotate axis labels
ggplot(data = pokedata, aes(x = fct_reorder(type_1, hp, median), y = hp)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.3, width = 0.2, color = "purple") + # alpha changes transparency, the smaller the more transparent 
  facet_wrap(~generation) + #separates into generation
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

That third plot (box plot plus `geom_jitter()`) is a small, cheap insurance policy against the Datasaurus problem. You get the summary and the raw points at the same time!

### Violin Plots!

```{r}
#| label: first violin plot
ggplot(data = pokedata, aes(x = type_1, y = hp)) +
  geom_violin(quantile.linetype = 1, quantiles = 0.5) +
  geom_boxplot(alpha = 0.5, width = 0.5)
#the shape of the line tells you where the most points are, helps you see the distribution
```

## Tip!

You can pipe `dplyr` and `ggplot2` together!

```{r, message=FALSE}
#| label: pipe dplyr with ggplot

# try a bar plot of type_2 first. see all the NAs?
ggplot(pokedata, aes(x = type_2)) +
  geom_bar()

# filter out NA values in type_2 and pipe directly to ggplot
pokedata |> 
  filter(!is.na(type_2)) |>  # filter out NA values, pipe directly to ggplot
  ggplot(aes(x = type_2)) + # ggplot still uses + as a layer operator
  geom_bar()

# another way to do this:
ggplot(na.omit(pokedata), aes(x = type_2)) +
  geom_bar()

```

## Challenge #1

Create a faceted bar plot that displays `type_1` and `type_2` information.

- Can you add color by `generation`?

- Can you rotate the labels on the x-axis? Make the font smaller?

```{r}
#| label: challenge 1: faceted stacked bar plot

ggplot(na.omit(pokedata), aes(x = type_2, fill = generation)) +
  geom_bar() +
  facet_wrap(~type_1) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1, size = 9))

```

## Covariation

Covariation refers to the tendency of two or more variables to change together in a correlated manner. A good way to inspect covariation is through visualization.

Below are plots using `geom_freqpoly()`, a function in that creates a frequency polygon, which is a line graph used to represent the distribution of a continuous variable.

Unlike histograms, which use bars, frequency polygons use lines, making it easier to visualize the shape of the distribution and compare multiple groups on the same plot. You can also overlay multiple frequency polygons to compare the distributions of different groups or categories, using different colors or line types.

Displaying counts is useful when you want to see absolute frequencies. Displaying density is helpful when you want to compare distributions across groups that don't have the same sample size. In the density display, the `y-axis` is scaled so that the area under the curve equals `1`.

**Counts** show the raw number of observations in each bin/category. This is useful if you care about absolute frequencies.

**Density** rescales those counts so the total area under the curve is 1. This standardization puts different groups on the same scale, which is is helpful when you want to compare distributions across groups that don't have the same sample size (you can compare their *shapes* for the distribution).

```{r, message=FALSE}
#| label: frequency polygon

# this one displays counts
ggplot(pokedata, aes(x = defense)) + 
  geom_freqpoly(aes(color = generation), binwidth = 15, linewidth = 0.75)

# this one displays density
ggplot(pokedata, aes(x = defense, y = after_stat(density))) + 
  geom_freqpoly(aes(color = generation), binwidth = 15, linewidth = 0.75)

```

## `diamonds` data

Here is the same sort of example as above, this time using data from `diamonds`, which is already part of your `ggplot2` package. Note that we can call upon and access the data without reading it in first!

```{r, message=FALSE}
#| label: examine diamonds data

ggplot(diamonds, aes(x = price)) + 
  geom_freqpoly(aes(color = cut), binwidth = 500, linewidth = 0.75)

ggplot(diamonds, aes(x = price, y = after_stat(density))) + 
  geom_freqpoly(aes(color = cut), binwidth = 500, linewidth = 0.75)

# and a boxplot for good measure
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot()

```

These data show that, contrary to expectations, better quality diamonds are perhaps generally cheaper than lower quality diamonds. This doesn't make sense. How do we dig deeper?

First, let's look at the different variables in the data

```{r, message=FALSE}
#| label: summarize diamonds data

head(diamonds)
summary(diamonds)
?diamonds

```

We can also summarize `price` by `cut` numerically, which makes the puzzle easier to state precisely:

```{r, message=FALSE}
#| label: descriptive summary of diamonds data

diamonds |> 
  group_by(cut) |> 
  summarise(
    n           = n(),
    mean_price  = mean(price),
    median_price = median(price))

```

Looks like there are other factors that could affect `price` – not just `cut`. Let's start with a basic scatter plot looking at `price` vs. `carat`.

```{r, message=FALSE}
#| label: scatter plot of carat x price

# try making the points transparent to 0.05
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05)

```

We can see the general trends, but it's a bit difficult to really see what is going on. Back to boxplots!

## Boxplots for EDA

First, we'll try making a boxplot with `x = carat` and `y = price`

```{r, message=FALSE}
#| label: box plot carat x price

ggplot(diamonds, aes(x = carat, y = price)) +
  geom_boxplot()

```

It doesn't work that way! Boxplots are designed for categorical (factor) variables on the x-axis, and `carat` is continuous.

But if you want to compare `price` across *ranges* of `carat`, you can do that by binning `carat` into increments of 1, 0.5, and 0.1 carats:

```{r, message=FALSE}
#| label: box plot carat x price (corrected with increments!)

ggplot(diamonds, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_width(carat, 1))) # note cut_width() 

ggplot(diamonds, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_width(carat, 0.5)))

ggplot(diamonds, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_width(carat, 0.1)))

```

We can see the general trend that an increase in diamond `carat` size results in an increase in `price`, with some outliers. We are getting closer to an answer. Let's see what `color` and `clarity` help us understand.

```{r, message=FALSE}
#| label: box plot color x price

ggplot(diamonds, aes(x = color, y = price)) +
  geom_boxplot()
# automatically plots the factor color in alphabetical order.

# FYI, color D (colorless) is considered better quality than color J (near colorless, more yellow), so let's reverse the order
diamonds |>
  mutate(color = fct_rev(color)) |> # fct_rev(color) reverses the order of factors
  ggplot(aes(x = color, y = price)) +
  geom_boxplot()

# color x carat now
diamonds |>
  mutate(color = fct_rev(color)) |> 
  ggplot(aes(x = color, y = carat)) +
  geom_boxplot()
```

Is there a relationship between `color` and `price`?

Let's try `clarity` (these are already in the "right" order).

```{r, message=FALSE}
#| label: box plot clarity

# clarity x price
ggplot(data = diamonds, aes(x = clarity, y = price)) +
  geom_boxplot()

# clarity x carat
ggplot(data = diamonds, aes(x = clarity, y = carat)) +
  geom_boxplot()

```

What is the relationship here?

For `color` and `clarity`, is there more visible variation or covariation?

Let's not forget about `cut`! Is there a relationship between `cut` and `carat`? `cut` is categorical and `carat` is numeric, and we can plot the two together easily

```{r, message=FALSE}
#| label: explore cut

# box plot of cut x price
ggplot(data = diamonds, aes(x = cut, y = price)) +
  geom_boxplot()

# box plot of cut x carat
ggplot(data = diamonds, aes(x = cut, y = carat)) +
  geom_boxplot()

# count plot
ggplot(data = diamonds, aes(x = cut, y = carat)) +
  geom_count()

# violin plot
ggplot(data = diamonds, aes(x = cut, y = carat)) +
  geom_violin()

```

Notice that the violin plot shows you something the boxplot hides: the actual shape of each distribution.

Between `color`, `clarity`, and `carat`, which do you think is the best predictor of `price`?

## Challenge #2:

### What do you think?

#### Observations:

- Where do we see a lot of variability?

- Is there a relationship between carat and cut?

- Which category has the largest carat diamonds?

#### Conclusions:

The variable with the most influence on price is \_\_\_.

### Reflection

1.  Which visualization type today felt most useful for you in spotting covariation, and why?\
    \
    Answer:

2.  If you had only been given the summary table for `diamonds` and no plots, which of today's conclusions could you have reached?\
    \
    Answer:
