---
title: "CS201 - Introduction & Getting Started with R & RStudio"
subtitle: "GitHub, warm-up, and a first look at mtcars"
author: "Amber Camp"
date: "2025-09-08"
format: html
editor: visual
---

## Welcome Back!

- Quick check-in: Who still needs help getting R, RStudio, and Git installed?
- Plan for today:
  - Revisit: R vs. RStudio
  - Git vs GitHub & why we care
  - Base R vs. tidyverse (conceptual only)
  - Play with `mtcars`

## R vs. RStudio

What is the difference between R and RStudio and why do you need to have both for this class?

- **R** is the programming language and engine
- **RStudio** is the IDE (Integrated Development Environment –your workspace) that makes using R easier
- You *can* run R without RStudio, but you probably wouldn’t want to (ask me how I know)

## Git vs. GitHub

Similarly, remember that Git and GitHub are *not the same thing*.

- **Git**: Tracks changes to your code locally on your computer
- **GitHub**: Hosts your Git repositories online for sharing, backup, and collaboration
- Together: You can track your work + share it with others

## Extra Credit Question (from last time)

Who was able to figure out why `0.1 + 0.1 + 0.1 - 0.3 ≠ 0`?

Try it!

```{r}
0.1 + 0.1 + 0.1 - 0.3
0.3 - 0.1 - 0.1 - 0.1 # a variation

# run this if you want to avoid scientific notation
options(scipen = 999)
```

Explanation:

- **Computers store numbers in binary (base-2)**, not decimal (base-10)

- Some decimal fractions — like `0.1` and `0.3` — **cannot be represented exactly** in binary

  - Think of how 1/3 can’t be written exactly as a finite decimal (0.3333… forever)

- When you add and subtract these approximations, you get a result that’s very close to zero, but not exactly

- This isn’t a bug! It’s how computers handle decimal math. The answer is *almost* zero, just not perfectly, because 0.1 can’t be stored exactly in binary

- You can check for "closeness" by using `all.equal(0.1 + 0.1 + 0.1, 0.3)` (should equal `TRUE`)

------------------------------------------------------------------------

# Let's get started!

## Install & Load packages

This gives you access to packages/libraries, which are like custom/specialized tools so that you can do more with R.

```{r, message = FALSE}
# install.packages("tidyverse")
library(tidyverse)
```

## Quick note on Base R vs. `tidyverse`

- **Base R**: built-in functions, available immediately
- **`tidyverse`**: collection of modern packages with consistent syntax
- **`tidyverse`** is considered the modern way to interact with R, and this will be our focus

### Example

Both of the following create a table from the dataset `mtcars` with columns for `mpg` and `hp`, filtering for 6 cylinder models and `mpg` \> 20.

**Base R:**

```{r, message = FALSE}
mtcars[mtcars$cyl == 6 & mtcars$mpg > 20, c("mpg", "hp")]
```

**tidyverse:**

```{r, message = FALSE}
mtcars %>%
  filter(cyl == 6, mpg > 20) |> # reminder that %>% is the same thing
  select(mpg, hp)
```

- Base R uses bracket notation and logical conditions
- `tidyverse` uses verbs like `filter()` and `select()` chained with pipes for readability
- Base R powerful and always available. `tidyverse` makes code more expressive and human-readable, especially for multi-step data wrangling (it's often easier to use!)

## First Look: `mtcars`

`mtcars` is a classic, built-in dataset. If you've learned R before, you've probably interacted with `mtcars`.

Start by exploring the data below. It's always good to ensure you understand what is contained in your dataset.

```{r, message = FALSE}
data(mtcars) # reads in the data & adds it to your environment as a dataframe
mtcars

```

- 

  ## How many rows and columns are in the dataframe?

- 

  ## What are the variables you see?

- 

  ## How might you view the last few rows of the dataframe with a single function?

### Quick Questions

- What is the average `mpg` overall?

```{r, warning=FALSE}


```

- Can you find the mean of another variable?

```{r, warning=FALSE}


```

- Can you find the median?

```{r, warning=FALSE}


```

### Quick Plot

Make a scatter plot of `mtcars` with `weight` on the x-axis and `mpg` on the y-axis. Use what you learned last class!

```{r, warning=FALSE}
# this is a scatterplot of mpg by wt


```

## Exit Ticket

Create a scatter plot and histogram of two variables of your choice (try to make it different from above). For your histogram, use `geom_histogram` and only assign an x variable in your `aes`.

Try beautifying your plots using what you have learned so far.

```{r, warning=FALSE}


```

Save, stage, commit, and push to GitHub!!
