---
title: "Treatment Mapping, part 2"
author: "Amber Camp" 
date: "`r Sys.Date()`"
format: html
editor: visual
toc: true
---

## Load Packages

If you don't have the below packages, run the relevant lines here:

```{r}
# only install these packages if you don't already have them!

# install.packages("tidyverse")
# install.packages("janitor")
# install.packages("sf")
# install.packages("remotes") # used to install packages from github
# remotes::install_github("UrbanInstitute/urbnmapr") # use this to install urbnmapr
# install.packages("naniar")
# install.packages("ggiraph")
```

Load packages (all of them).

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor) 
library(sf) # "simple features" for spatial data
library(urbnmapr) # access to US state/country shapefiles for mapping
library(naniar) # tools for exploring and visualizing missing data
library(ggiraph) # makes ggplot charts interactive
options(scipen = 99) # turns off scientific notation
```

## Today's Data

Read in the data you cleaned on Monday! [**THIS WILL NOT WORK IF YOU WEREN'T HERE ON MONDAY!**]{.underline}

As a reminder, the data is from the 2024 Treatment Episode Data Set-Discharges (TEDS-D) Dataset from the Substance Abuse and Mental Health Services Administration (SAMHSA). The metadata codebook can be found [here](https://www.samhsa.gov/data/system/files/media-puf-file/TEDS-D-2024-DS0001-info-codebook_v1_0.pdf)

```{r, echo=FALSE}
teds_d <- readRDS("data/teds_d_clean.rds")
```

### Columns for this data

- `freq1_d` = Frequency of use at discharge

- `stfips` = State FIPS code (FIPS = *Federal Information Processing Standard)*

- `services_d` = Treatment Service

- `los` = Length of Stay

- `reason` = Reason for Discharge

## Mapping the Data!

Get ready for some more "word problems." A big part of this class is translating regular language to code-based solutions, and vice-versa.

Our first goal is to map the [*percentage of complete treatments by state*]{.underline}*.*\
Reworded: *What percentage of treatment episodes in each state ended in completion?*

We'll break it down. First, let's calculate the percentage of completed treatments by state.

```{r}
#| label: percent-completed-by-state

# Fill in the blanks. Ask yourself: what am I counting, and out of what?

percent_completed_by_state <- teds_d |>
  group_by(stfips) |> 
  summarize(total_cases = n(),
            completed_cases = sum(reason == "Treatment completed", na.rm = TRUE)) |> 
  mutate(percentage_completed = (completed_cases / total_cases) * 100)

summary(percent_completed_by_state)
```

### Check your work

Before moving on, sanity-check what you just made. Good data scientists check their numbers *before* they make something pretty out of them.

- Are all of your percentages between 0 and 100?

- Nationally, 47.8% of 2024 discharges ended in completion (check the codebook). Are your numbers in that neighborhood?

- Does `sum(percent_completed_by_state$total_cases)` equal `nrow(teds_d)`?

> **Answers:**

```{r}
#| label: check-percentages

summary(percent_completed_by_state$percentage_completed)

sum(percent_completed_by_state$total_cases) == nrow(teds_d)

```

## Bringing in map data

Next, we need to bring in some mapping data using `urbnmapr`. Notice there is no map data in our current dataset, so we wouldn't be able to create maps without the next few steps.

```{r}
#| label: get-map-data
#| message: false
#| warning: false

states_map <- get_urbn_map(map = "states", sf = TRUE)

head(states_map, 10)
```

Take a look at what came back. What is in the `geometry` column?

> **Answer: shapes the states**

### Finding an index

Next, we will want to combine the `states_map` and `percent_completed_by_state` dataframes. In order to do that, we need to find something they have in common — an `index`.

That index will be the state FIPS code. But what do we notice that's different between the columns holding FIPS information in `percent_completed_by_state` and in `states_map`? (Two differences.)

> **Answer:** (1) The columns are named differently: `stfips` in ours, `state_fips` in theirs. (2) Ours are numbers (1, 2, 4...) while theirs are text with a leading zero ("01", "02", "04"...).

So we need to fix both differences before the two tables can talk to each other.

```{r}
#| label: prepare-index

# copies stfips data to create new column in percent_completed_by_state called 
# state_fips, forcing values to be two digits (start with zero)

percent_completed_by_state <- percent_completed_by_state |>
  mutate(state_fips = sprintf("%02d", stfips)) # sprintf = string print formatted

head(percent_completed_by_state)
```

`sprintf("%02d", stfips)` reads as: "print this number as text, using at least 2 digits, padding with a zero if needed."

Why do FIPS codes have to be text instead of numbers?

> **Answer:**

### Joining data

Here is a \~\*magical\*\~ concept in data science: ***joins***! They let us connect information that lives in different tables, like linking student names in one list with their grades in another... or linking treatment discharge data with state mapping data based on an index of `state_fips`!

This is both a useful **concept** (relating tables through keys) and a useful **skill** (choosing and applying the right type of join), carried out in R with functions like `left_join()`, `right_join()`, or `full_join()`.

```{r}
#| label: join-data
#| message: false
#| warning: false

mapping_data <- full_join(percent_completed_by_state,
                          states_map,
                          by = "state_fips") # remember the index we prepared above?

```

Success! This was easily done with a single line of code—but don't underestimate the power of what you just did, and how precise things needed to be for you to make it work.

But now compare the number of rows. `percent_completed_by_state` has one row per state *in our data*. `states_map` has one row per state *on the map*. How many rows does the joined version have, and why isn't it the same as either one?

> **Answer:**

```{r}
#| label: compare-row-counts

nrow(percent_completed_by_state)
nrow(states_map)
nrow(mapping_data)
```

### Plotting the Map

```{r}
#| label: first-map

ggplot(mapping_data) +
  geom_sf(mapping = aes(geometry = geometry, fill = percentage_completed), color = "#ffffff", linewidth = 1) +
  labs(fill = "% of Completed Treatment Episodes") +
   coord_sf(datum = NA) + # turns off lat/long grid
   theme_minimal() 
```

Make it interactive with `ggiraph`. This package works well with `ggplot2`.

```{r}
#| label: first-interactive-map

completed_treatment_map_int <- ggplot(mapping_data) +
  geom_sf_interactive(mapping = aes(geometry = geometry, fill = percentage_completed,
      tooltip = paste("State FIPS:", stfips, 
                      "Completed:", percentage_completed, "%")),
    color = "#ffffff",
    linewidth = 0.25) + #changes the white outline
  labs(fill = "% of Completed Treatment Episodes") +
  coord_sf(datum = NA) + # turns off lat/long grid
  theme_minimal()


# Use `girafe` to render the interactive plot
girafe(ggobj = completed_treatment_map_int) # check your spelling!

```

### Clean up the tooltip

You can customize the tooltip (to a degree). Hover over a few states and notice two annoyances: the percentage runs to many decimal places, and the tooltip identifies states by a number nobody has memorized.

Try rounding that completed % value to 2 decimal points and adding the state name. Which column in your joined data holds the state name or abbreviation?

> **Answer:**

```{r}
#| label: tooltip-cleanup

completed_treatment_map_int <- ggplot(mapping_data) +
  geom_sf_interactive(mapping = aes(geometry = geometry, fill = percentage_completed,
                                    tooltip = paste("State Abbreviation:", state_abbv,
                                      "<br>State FIPS:", stfips, 
                                                    "<br>Completed:", round(percentage_completed,2), "%")),
                      color = "#ffffff",
                      linewidth = 0.25) + 
  labs(fill = "% of Completed Treatment Episodes") +
  coord_sf(datum = NA) + 
  theme_minimal()


# Use `girafe` to render the interactive plot
girafe(ggobj = completed_treatment_map_int) 

```

### Clean up the color binning

Right now our fill is a continuous gradient, which makes it hard to tell similar states apart. Let's group the percentages into bins instead.

Adding color bins, based on manual/equal interval breaks. (There are more advanced binning techniques, like quantiles, Jenks natural breaks, standard deviations, etc.)

```{r}
#| label: add-bins

mapping_data <- mapping_data |> 
  mutate(percentage_bin = cut(percentage_completed, breaks=c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100)))

# check out the column you just added
head(mapping_data)

```

`cut()` reads as: "take this numeric column and sort each value into one of these labeled ranges." What does a bin label like `(40,50]` mean?

> **Answer:**

Now make the map!

```{r}
#| label: binned-map

percent_complete_binned_int <- ggplot(mapping_data) +
  geom_sf_interactive(
    aes(geometry = geometry,
      fill = percentage_bin,
      tooltip = paste("State FIPS:", stfips, # add state abbreviation if you want
                      "<br>Completed:", round(percentage_completed,2), "%")),
    color = "#ffffff", linewidth = 0.25) +
  labs(fill = "% of Completed Treatment Episodes",
       title = "Completed Treatment Episodes by State",
       subtitle = "TEDS-D 2024 (SAMHSA)") +
  scale_fill_viridis_d(option = "mako") + 
  coord_sf(datum = NA) +
  theme_minimal() +
  theme(panel.background = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.text = element_text(size = 4),
    legend.title = element_text(size = 5),
    strip.text = element_text(size = 4))

girafe(ggobj = percent_complete_binned_int)
```

## For Friday's Class

Notice anything weird with the above?? Try both to identify the problem and to figure out how to fix it—you'll need it for your homework. I'll show you a solution on Friday.

### Notes:

### Sandbox

```{r}

```
